---
title: "STS 101 Lab 2"
author: ""
date: "October 10, 2025"
output: html_notebook
---
<body style="background-color:black; color:white">

Begin by sweeping your environment and restarting R.

## Getting started

Load the `tidyverse` package, read in the `kids-subset.csv` and `parents-subset.csv` data frames as we did in `Notebook 2`, then join them into a data frame called `joined`. In this data frame, you want all children and the parents that link to them.
```{r}
library(tidyverse)

kids <- read.csv("kids-subset.csv") |>
        select(-X)
parents <- read.csv("parents-subset.csv") |>
        select(-X)

joined <- inner_join(kids, parents, by = "PARID")
```

## Parents' ages at marriage (2 points)

Using the `year_span()` function we made in `Notebook 1`, create a variable called `MOM_AOM` that calculates the mother's age at marriage (in years) and a variable called `DAD_AOM` that calculates the father's age at marriage (in years). *Make these new variables in the `parents` data frame.* Use the `summary()` function to examine these new variables.
```{r}
#Read in year_span() function
source("year_span.r")

#Create new variables
parents <- joined |> 
              mutate(MOM_AOM = year_span(MOM_DOB,DOM),
                     DAD_AOM = year_span(DAD_DOB, DOM))

#Examine new variables
summary(parents$MOM_AOM, rm.na = F)
summary(parents$DAD_AOM, rm.na = F)
```
What is the likeliest cause of the implausible values you see here?

Ans: These implausible values might because typo(they wrote wrong, for example, they wrote 1660, but in reality it's 1616) or they wrote in wrong place when they are recording the data.
> 


Graph the distribution of the parents' ages at marriage side by side. Include only ages from 0 to 100. Do this first as a histogram and then as a density plot. You may use either facets or `ggarrange()`. Remember to include appropriate labels and use a theme other than the default.
```{r}
#Histogram 
parents |> 
  select(KIDID, MOM_AOM, DAD_AOM) |>
  pivot_longer(!KIDID, names_to = "parent", values_to = "age") |>
  filter(!is.na(age)) |>
  mutate(parent = str_sub(parent, 1, 3)) |>
  ggplot(aes(age)) +
  geom_histogram(binwidth = 1) +
  theme_bw() +
  scale_x_continuous(lim = c(0, 100), breaks = seq(0, 100, 10)) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Age", y = "Number of people who married",
       title = "Age of marriage recorded in Henry data") +
  facet_wrap(~parent, ncol = 2) #combine together

#Density plot
parents |> 
  select(KIDID, MOM_AOM, DAD_AOM) |>
  pivot_longer(!KIDID, names_to = "parent", values_to = "age") |>
  filter(!is.na(age)) |>
  mutate(parent = str_sub(parent, 1, 3)) |>
  ggplot(aes(age)) +
  geom_density() +
  theme_bw() +
  scale_x_continuous(lim = c(0, 100), breaks = seq(0, 100, 10)) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Age", y = "percent of people who married",
       title = "Age of marriage recorded in Henry data") +
  facet_wrap(~parent, ncol = 2)
```



Again using the `year_span()` function, create a variable in the `parents` data frame called `AGE_DIFF` that represents the span in years from the father's date of birth to the mother's date of birth (the age difference between the parents). Use the `summary()` function to examine this variable.

```{r}
#Create new variable
parents <- joined |> 
              mutate(AGE_DIFF = year_span(DAD_DOB, MOM_DOB)) # MOM'age-DAD'age
#Examine new variable
summary(parents$AGE_DIFF, rm.na = F)
```

A positive number indicates that which spouse is older?

Ans: A positive number indicates that dad is older than mom in this family.
> 

Create a histogram showing the age difference between each couple. Use a `binwidth` of 1 and include only couples with an age difference of no more than 25 years in either direction. Include a title and axis labels, and change the theme to something other than the default. A viewer should be able to tell from your title and/or axis labels what a positive number indicates.
```{r}
parents |> 
  ggplot(aes(AGE_DIFF)) +
  geom_histogram(binwidth = 1) +
  theme_bw() +
  scale_x_continuous(lim = c(-25,25), breaks = seq(-25, 25, 5)) +
  labs(x = "Age Difference(Positive means DAD is older than MOM in this couple)", y = "Number of couples",
       title = "Age difference of couples in Henry data set")
```

## Time from marriage to first birth (2 points)

In the notebook, we considered the age of mothers at the birth of their *last* child. Now we want to know how soon after marriage a couple's *first* child was born.  Create a data frame called `oldest` that includes only the rows for each couple's oldest child from the `joined` data frame. Remember that in the notebook, we used the `max()` function to find the child born most recently. Now we need to use the `min()` function to find the child born longest ago. Once we find it, we can use the `interval()` function and the `months()` function to create a variable called `MONTHS` that indicates how many months after marriage the child was born. Use the `summary()` function to examine the new variable.
```{r}
#Create the new dataset and variable; remember to keep date of marriage by including it in group_by()
oldest <- joined |>
            group_by(PARID, DOM) |> #we want keep parid and dom
            summarize(DOB = min(as_date(DOB))) |>
            mutate(MONTHS = interval(DOM, DOB) / months(1))

#Examine the new variable
summary(oldest$MONTHS)
          
```
Create a histogram (with binwidth representing one month) and density plot showing the number of months between marriage and first birth for each couple, including only values between 0 and 48 (the first four years of marriage).
```{r}
#Histogram 
#BINWIDTH 1, gap 4 yrs
oldest |>
  ggplot(aes(MONTHS)) +
  geom_histogram(binwidth = 1) +
  theme_bw() +
  scale_x_continuous(lim = c(0, 48), breaks = seq(0, 48, 2)) +
  labs(x = "Months of mothers at first birth from marriage", y = "Number of mothers",
       title = "Months of mothers at first birth in Henry data set")
  


#Density plot
oldest |>
  ggplot(aes(MONTHS)) +
  geom_density() +
  theme_bw() +
  scale_x_continuous(lim = c(0, 48), breaks = seq(0, 48, 2)) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Months of mothers at first birth from marriage", y = "Percent of mothers",
       title = "Months of mothers at first birth in Henry data set")

```






